Splitting it Up: The \pkg{spduration} Split-population Duration Regression Package

### 2016-03-07 manuscript.tex
how about actually syncing manuscript2.tex
also expanded alternative implementation section

\author{
Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting it up: the \pkg{spduration} split-population duration
regression package}
\Keywords{keywords, not capitalized, \proglang{R}}

\Abstract{
We present an implementation of split-population duration regression in
the \pkg{spduration} R package and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration model. The package includes estimation and forecasting
methods for split-population Weibull and Loglogistic models.
}

\Plainauthor{Daniel W. Hill, Nils W. Metternich, Andreas Beger, Michael D. Ward}
\Plaintitle{Splitting it up: the spduration split-population duration regression
package}
\Shorttitle{\pkg{spduration}: Split-population duration regression}
\Plainkeywords{keywords, not capitalized, R}

@@

\Address{
    Daniel W. Hill\\
  University of Georgia\\
  First line Second line\\
  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
  URL: \url{http://spia.uga.edu/directory/faculty_staff/hill-daniel}\\~\\
        }

\usepackage{amsmath}

@@

This template demonstrates some of the basic latex you'll need to know
to create a JSS article.

\subsection{Code formatting}\label{code-formatting}

Don't use markdown, instead use the more precise latex commands:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \proglang{Java}
\item
  \pkg{plyr}
\item
  \code{print("abc")}
\end{itemize}

\section{Introduction}\label{introduction-1}

Duration models are an important class of statistical estimators that
take into account the duration dependency of outcomes we are interested
in. For example, the risk of dying is not independent of time. Newborns
are at a great risk of dying, but as they grow older this risk declines
and then gradually starts to increase again after the age of 9-10. This
risk can be increased or decreased by behavioral (smoking, exercise,
diet, etc) and structural factors (health care, regulations, urban vs
rural, etc), but there is an underlying risk that is time dependent and
impacts on all units.

However, sometimes we are interested in situations in which not all
units are at risk of dying, failing, or acquiring some new
characteristic. Imagine we are interested in the risk of acquiring the
flu. First, we might think that everyone is at risk of getting the flu
and it just depends on individual behavior (e.g.~good hygiene) and
structural factors (e.g.~workplace) whether individuals will acquire the
flu. If this would be the case we could use a standard duration model
and estimate how different factors impact on the baseline risk. But
there might also be individuals who are immune to the flu, because they
received a flu shot, had the same flu virus in the past year, or some
other factors that makes it impossible for them to get sick. Given that
this is the case we have two underlying populations: An at risk
population and an immune one. If the immune population is relatively
large, estimates using a standard duration model will be biased and
predictions resulting from such a model inaccurate.

@@

Duration models, where the variable to be modeled describes a
distribution of times until some event of interest, were originally
developed in health and demographic research, where they grew naturally
from life tables and survival records for medical patients. Basic
formulations of such models, like the parametric exponential or Weibull
regressions or semi-paramaetric Cox regression, implicitly assume that
all subjects or units under observation, including right-censored
observations, will eventually experience the event of interest. This
assumption is outright wrong or at least disadvantageous in many
substantive areas and empirical applications, where there instead is a
sub-population of units or subjects that will never experience an event,
and thus are effectively ``cured''.

This insight was realized as early as 1949 by Boag (1949) and Berkson
and Gage (1952), who were researching survival rates in cancer patients
following treatment, and where quite obviously some fraction of patients
survived because their cancer was cured, while others relapsed after
apparent remission due to levels of disease below detectable thresholds.
As with conventional duration models, their origin in health and
medicine has shaped the terminology conventionally used with such model,
e.g.~split-population duration models are sometimes also referred to in
such contexts as cure rate models, and the basic concepts like survival
and failure rates reference the survival of humans.

Yet the intuition underlying split-population duration models has led to
applications in a broad range of subject areas outside demographics and
medicine. In an early and foundational application that reached beyond
criminal science, Schmidt and Witte (1989) examined criminal recidivism
using data on close to 10,000 prisoners from the North Carolina prison
system in the late 1970's and early 1980's to identify factors that
influence whether a criminal lapses at all, and if so which factors are
related to the amount of time between prison stints. This work already
includes a full formulation of the model with independent covariates for
both the duration equation and the risk or cure equation, although only
with subject-specific covariates rather than time-varying covariates
with multiple data points per subject.

In public health, Douglas and Hariharan (1994) use data from the US to
model factors related to the age at which smokers started the habit, and
Forster and Jones (2001) examines the impact of tobacco taxes on smoking
and quitting decisions. DeYoung (2003) models the failure of new
commercial banks in the US during the 1980's. Lastly, in our own domain,
political science, Svolik (2008) used a split-population duration
framework to study democratic consolidation, i.e.~whether democratic
regimes persist or slide back to authoritarianism, and when. Building on
this effort, split-population duration models have also been used, along
with other models, to produce regular predictions for five different
forms of political conflict for the Integrated Crisis and Early Warning
System (ICEWS) project (M. D. Ward et al. 2013) and to model irregular
leadership changes for the Political Instability Task Force (PITF;
Beger, Dorff, and Ward 2014).

@@

Conventional duration models assume that all subjects will eventually
fail, if we can only observe them long enough. The likelihood for a data
point with survival time \(t\) is thus the failure rate at that time or
the probability of survival beyond \(t\), depending on whether the
observation is right-censored (\(1-\delta_i\)) or not:

@@

The major modeling question in this setting, which we will return to
below, is the choice of a function to describe the evolution of the
hazard rate \(h(t) = \frac{f(t)}{S(t)}\) over time, e.g.~with
exponential, Weibull, or log-logistic densities.

The cumulative failure rate (\(F(t) = 1 - S(t)\)) over time converges to
1, meaning all subjects fail eventually. This assumption is untenable in
many situations. Some cancer patients are cured after treatment while
others relapse, most young people do not start smoking even though most
smokers started at a young age, and many states will not experience the
kind of violence that persists in some parts of the world. A better
assumption in such contexts is that there is a subpopulation of subjects
who are at risk of experiencing an event at some point, and another
subpopulation who will never experience the event.

The presence of a large sub-population which is not at risk for an event
in practice will inflate estimates of the survival fraction, and reduce
hazard estimates for all subjects by essentially, and falsely,
distributing risk among subjects that genuinely will fail and those that
are cured. A model will thus over predict hazard for subjects that are
not at risk (cured), and under predict for those who are at risk of
eventually experiencing the event of interest.

@@

Crucially, this split-population framework is primarily useful in
contexts where sub-populations are not clearly or easily identifiable.
Even though there is a clear sub-populations in a model of the age at
first pregnancy for humans---men---no researcher in their right mind
would think to include male subjects in their data at all, for example.
On the other hand, whether a cancer patient is cured or not cured given
that they have no visible signs of cancer following treatment or have
hit the 5-year disease free survival mark is much less clear and in fact
the question that initially led to the idea for split-population
duration modeling (Boag 1949, Berkson and Gage (1952)).

@@

Where \(z_i\) is a vector of covariates for a subject at a given time.
For interpretation, it is important to note that with time-varying
covariates, the risk (or cured) estimate for a subject is particular to
a given time point rather than constant for that subject over all time
periods in the
spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time. Subjects can theoretically have multiple spells, e.g. cancer patients who go into remission and relapse more than once, or states that experience multiple civil war onsets over their history.}
Depending on the covariates, the risk estimate for a subject can thus
fluctuate rapidly over time. To ease interpretation, it might be
convenient to restrict covariates in the logit risk model to
slow-moving, stable covariates in order to produce stable risk estimates
for subjects.

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \pkg{spduration}
package implements two hazard rate shapes,
Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.}
and log-logistic:

@@

Where \(\lambda = e^{-x_i\beta}\) is a parameter of covariates.

With these, the main quantity of interest is the conditional hazard
\(h(t, \pi)\), where both the risk/cure probabilities and hazard are
conditional on survival to time \(t\):

@@

For a given unconditional risk rate \(\pi\), the probability that a
particular case with survival time \(t\) is in the risk set decreases
over time because an increasing number of surviving cases consist of
immune or cured (\(1-\pi\)) cases that will never fail. In the hazard
rate, the failure rate in the numerator is conditional on the
probability that a case is in the risk set, give survival up to time
\(t\), and the numerator is an adjusted survivor function that accounts
for the fraction of cured cases by time \(t\), which is \(1-\pi(t)\).

@@

The \pkg{spduration} package for \proglang{R} implements a
split-population duration regression for data with time-varying
covariates and with two options for the hazard rate shape, Weibull and
log-logistic. The Weibull density allows for hazard rates that are
increasing, constant, or decreasing over survival time, while the
log-logistic density also can fit rates that have a peak at a particular
survival time.

As an example to illustrate the package functionality, we will replicate
and extend the model of coup onsets in Belkin and Schofer (2003). Belkin
and Schofer's paper lends itself to re-analysis with a split-populuation
duration model because they explicitly distinguish long-term structural
risk factors for coups from more short-term triggering causes that can
explain the timing of a coup in a regime at risk. They argue that many
countries never experience coups because coups are effectvely impossible
due to structural factors, while others that never experience coups are
nevertheless at risk due to a different configuration of those same
factors. Using language which fits nicely with the class of models
described above, they say ``Triggers are not the source of the original
risk, and in the absence of structural causes, the presence of
triggering factors alone cannot lead to a coup. Hence, triggers should
not be equated with coup risk. Rather, they are factors that may
determine the exact timing of a coup in regimes that suffer from high
coup risk'' (p.~598 Belkin and Schofer 2003)

Belkin and Schofer develop a measure of ``structural coup risk'' which
incorporates factors such as the strength of democratic institutions and
civil society, and a recent history of successful coups. They employ a
logistic regression model which does not adequately capture the process
described in the quote above, since it implicitly assumes that all
observations are at risk for a coup (i.e., the probability of a coup is
non-zero for all observations). Their structural coup risk indicator is
developed precisely to distinguish between cases where coups are
possible and cases where they are not, so the split-population model
allows one to examine whether the indicator effectively separates
at-risk cases from cases where coups are impossible.

@@

\begin{CodeChunk}
\begin{CodeInput}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

\normalsize
The \code{add\_duration} function takes as input a data frame with a
binary response variable (in this example, ``coup''``) that measures the
occurrence of the event, or failure, recorded over discrete time
periods, and produces a data frame with a duration variable
(named''duration``) for each spell as well as a binary variable
(''atrisk``) that is constant across the spell and indicates whether the
spell ended in failure. These are then used as response variables for
the formula in the \texttt{spdur} function. The \code{spdur} function is
the primary function in the package and produces a regression model
object of class \code{spdur} which can then be used with further
methods.

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data. We begin by fitting first a
Weibull and then a log-logistic split-population duration model using
the coups data, including the measure of coup risk in the logit (risk)
equation.

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. One may also
use the \code{AIC} and \code{BIC} function to calculate the information
criterion statistics for \code{spdur} objects.

\begin{CodeChunk}
\begin{CodeInput}
AIC(weib_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1207.992
\end{CodeOutput}
\begin{CodeInput}
AIC(loglog_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1110.829
\end{CodeOutput}
\begin{CodeInput}
BIC(weib_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1290.452
\end{CodeOutput}
\begin{CodeInput}
BIC(loglog_model)
\end{CodeInput}
\begin{CodeOutput}
[1] 1193.289
\end{CodeOutput}
\end{CodeChunk}

The objects produced by \code{spdur} can also be used with functions in
the \code{xtable} package to produce \LaTeX~code for tables of results.
The \code{xtable} function used on an \code{spdur} object will produce
code such as that used to create Table \ref{loglog_table}.

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 3.30 & 0.14 & 0.00 \\ 
  milreg & -1.53 & 0.15 & 0.00 \\ 
  instab & -0.22 & 0.03 & 0.00 \\ 
  regconf & 3.04 & 2.22 & 0.17 \\ 
  log(alpha) & -0.52 & 0.07 & 0.00 \\ 
  (Risk Intercept) & -14.61 & 9.02 & 0.11 \\ 
  couprisk & 9.11 & 4.37 & 0.04 \\ 
  wealth & 3.51 & 1.75 & 0.05 \\ 
  milreg.1 & 21.76 & 0.24 & 0.00 \\ 
  rwar & -6.37 & 5.38 & 0.24 \\ 
  regconf.1 & -20.22 & 54.04 & 0.71 \\ 
  samerica & -3.31 & 3.28 & 0.31 \\ 
  camerica & -6.46 & 2.57 & 0.01 \\ 
   \hline
\end{tabular}
\end{table}

The table shows estimates from the duration equation, beginning with the
intercept, and then estimates from the risk equation. Interestingly, the
coup risk variable has a rather large and statistically significant
coefficient in the risk equation.

@@

\begin{CodeChunk}
\begin{CodeInput}
plot(loglog_model, type="hazard", main = "A")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
plot(loglog_model, type="hazard", 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
\end{CodeInput}
\end{CodeChunk}

Importantly, the package also includes functions to evaluate model
predictions. The generic function \code{predict} can be used on an
object of class \code{spdur} to generate several kinds of predictions,
including the probability that an observation is ``at-risk'' and the
probability of failure for a given time period. In addition, the generic
\code{plot} function can be used on \code{spdur} objects to produce a
separation plot (Greenhill, Ward, and Sacks 2011), which is a graphical
display for evaluating model predictions. The code below produces Figure
\ref{insamp}.

@@

Finally, we demonstrate how to evaluate a model's out of sample
predictions. We begin by splitting the data into training and test sets,
using the last year of data (1999) as the test set and omitting
incomplete observations from the test set.

\begin{CodeChunk}
\begin{CodeInput}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year")
\end{CodeInput}
\begin{CodeOutput}
Warning in attempt_date(data[, tID], freq): Converting to 'Date' class with
yyyy-06-30
\end{CodeOutput}
\begin{CodeInput}
coup_test  <- coup_test[coup_test$year >= 1996, ]
\end{CodeInput}
\end{CodeChunk}

\normalsize
The \code{spdurCrisp} function can then be used to fit a model on the
training set and generate predictions for the test set.

>>>>

\author{
Andreas Beger\\Duke University \And Daniel W. Hill, Jr.\\University of Georgia \AND Nils W. Metternich\\University College London \And Shahryar Minhas\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting it Up: The \pkg{spduration} Split-population Duration
Regression Package}
\Keywords{regression, rare events, \proglang{R}}

\Abstract{
We present an implementation of split-population duration regression in
the \pkg{spduration} R package and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models.
}

\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
\Plaintitle{Splitting it up: the spduration split-population duration regression
package}
\Shorttitle{\pkg{spduration}: Split-population duration regression}
\Plainkeywords{regression, rare events, R}

@@

\Address{
    Andreas Beger\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:ab428@duke.edu}{\nolinkurl{ab428@duke.edu}}\\
  URL: something\\~\\
      Daniel W. Hill, Jr.\\
  University of Georgia\\
  Department of International Affairs, University of Georgia Athens, GA,
  USA\\
  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\

      Nils W. Metternich\\
  University College London\\
  Department of Political Science, University College London London, UK\\
  E-mail: \href{mailto:n.metternich@ucl.ac.uk}{\nolinkurl{n.metternich@ucl.ac.uk}}\\

      Shahryar Minhas\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:sfm12@duke.edu}{\nolinkurl{sfm12@duke.edu}}\\

      Michael D. Ward\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:mw160@duke.edu}{\nolinkurl{mw160@duke.edu}}\\

  }

\usepackage{amsmath} \usepackage{booktabs}

@@

Duration models are an important class of statistical estimators that
take into account the duration dependency of specific outcomes. A
prominent example is that the risk of dying depends on the age of an
individual. Newborns are at a greater risk of dying, but as they grow
older this risk quickly declines and then gradually starts to increase
again after the age of 9-10. While individual behavior (smoking,
exercising, diet) and structural factors (health care, health
regulations, urban vs rural) can increase or decrease the probability of
dying over time, the underlying risk is time dependent and impacts on
all individuals.

However, there are conditions under which not all individuals have the
same underlying risk to experience a specific outcome and might not even
be at risk at all. Consider the risk of acquiring a viral infection like
the common flu. Let us initially assume that everyone is at risk of
infection, and individual behavior (e.g., good hygiene) and structural
factors (e.g., workplace) determine whether individuals catch the flu.
Under these assumptions a standard duration model should provide us with
efficient estimates of how different behavioral and structural factors
impact on the general baseline risk. But, there might be individuals
that are immune to the flu in the overall population--because they
received vaccination, had the specific flu virus in the past, or have
some other characteristic that makes it impossible for them to attract
the disease. In this instance we have two underlying populations: an at
risk population and an immune one. If the immune population is
relatively large, estimates using a standard duration model will be
biased and predictions inaccurate.

@@

Regular duration models, where baseline risk is modeled by some
distribution of time, were originally developed in health and
demographic research and grew naturally from life tables and survival
records for medical patients. Basic formulations of such models, like
the parametric exponential or Weibull regressions or semi-parametric Cox
regression, implicitly assume that all subjects or units under
observation, including right-censored observations, will eventually
experience the event of interest. This assumption may be violated in
many substantive areas and empirical applications, where a
sub-population of units or individuals will never experience an event,
and thus are effectively ``cured''.

The insight that populations might be split in regard to their baseline
risk has been formulated as early as 1949 by Boag (1949) and Berkson and
Gage (1952), who were researching survival rates in cancer patients. In
their application some fraction of patients survived because their
cancer was cured, while others relapsed after apparent remission due to
levels of disease below detectable thresholds. The development of
duration methods in health and medicine has also shaped the terminology
conventionally used for such models. For example, split-population
duration models are also referred to as cure rate models. Similarly, the
basic concepts like survival and failure rates reference the survival of
humans.

Yet the intuition underlying split-population duration models has led to
applications in a broad range of subject areas outside demographics and
medicine. In an early and foundational application that reached beyond
medicine, Schmidt and Witte (1989) examined criminal recidivism using
data on close to 10,000 prisoners from the North Carolina prison system
in the late 1970's and early 1980's to identify factors that influence
whether a criminal relapses at all, and if so which factors are related
to the amount of time between prison stints. This work already includes
a full formulation of a model with independent covariates for both the
duration equation and the risk or cure equation, although only with
subject-specific covariates rather than time-varying covariates for
multiple data points per individual.

In a complementary set of research from public health, Douglas and
Hariharan (1994) use data from the United States to model the age at
which individuals began to smoke, and Forster and Jones (2001) examine
the impact of tobacco taxes on smoking and quitting decisions. DeYoung
(2003), an economist, models the failure of new commercial banks in the
US during the 1980's. Svolik (2008) uses a split-population duration
framework to examine whether democratic regimes persist or revert to
authoritarianism, and when. Building on this effort, split-population
duration models have also been used, along with other models, to produce
regular predictions for five different forms of political conflict for
the Integrated Crisis and Early Warning System (ICEWS) project (Ward et
al. 2013) and to model irregular leadership changes for the Political
Instability Task Force (PITF; Beger, Dorff, and Ward 2014).

@@

Conventional duration models assume that all subjects will eventually
fail, die, or experience a specific outcome. The likelihood for a data
point with survival time \(t\) is thus the failure rate \(f(t)\) at that
time or the probability of survival beyond \(t\), \(S(t)\), depending on
whether the subject has already experienced the event (\(\delta_i\)) or
is right-censored (\(1-\delta_i\)):

@@

The major modeling question in this setting, which we will return to
below, is the choice of a function form (e.g., exponential, Weibull, or
log-logistic) describing the underlying hazard rate
\(h(t) = \frac{f(t)}{S(t)}\) over time.

The cumulative failure rate (\(F(t) = 1 - S(t)\)) over time converges to
1, meaning all subjects fail eventually. In the examples of applied
research discussed above this assumption is untenable. Some cancer
patients are cured after treatment, most young people never become
regular smokers, and many states will not experience the kind of
violence that persists in other parts of the world. The presence of a
large sub-population which is not at risk for an event, will in practice
inflate estimates of the survival fraction, and reduce hazard estimates
for all subjects. This is the case because the underlying risk is
estimated based on subjects that genuinely will fail and those that are
cured. Hence, such a model will over predict the hazard for subjects
that are not at risk (cured), and under predict for those who are at
risk of eventually experiencing the event of interest.

@@

Crucially, this split-population framework is primarily useful in
contexts where sub-populations are not clearly or easily identifiable.
For example, there is a clear sub-population in a model of the age at
first pregnancy for humans---men---which researchers can easily identify
and exclude from their data. Less clear are answers to questions such as
whether a cancer patient is cured or not cured given that they have no
visible signs of cancer following treatment or have hit the 5-year
disease free survival mark. In such situations split-population models
provide a way to infer sub-populations in a probabilistic fashion.

@@

Where \(z_i\) is a vector of covariates for a subject at a given time.
For interpretation, it is important to note that with time-varying
covariates, the risk (or cured) estimate for a subject is particular to
a given time point rather than constant over all time periods in the
spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time. Subjects can theoretically have multiple spells, e.g., cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.}
Depending on the covariates, the risk estimate for a subject can thus
fluctuate over time. To ease interpretation, it might be convenient to
restrict covariates in the logit risk model to slow-moving, stable
covariates in order to produce stable risk estimates for subjects.

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \pkg{spduration}
package implements two hazard rate shapes, Weibull and log-logistic:

@@

Where \(\lambda = e^{-x_i\beta}\) is a parameter of covariates. The
Weibull density allows for hazard rates that are increasing, constant,
or decreasing over survival time, while the log-logistic density also
can fit rates that have a peak at a particular survival time.

Given the density distribution, the main quantity of interest is the
conditional hazard \(h(t, \pi)\), where both the risk/cure probabilities
and hazard are conditional on survival to time \(t\):

@@

For a given unconditional risk rate \(\pi\), the probability that a
subject with survival time \(t\) is in the risk set decreases over time,
because an increasing number of surviving cases consist of immune or
cured (\(1-\pi\)) cases that will never fail. In the hazard rate, the
failure rate in the numerator is conditional on the probability that a
case is in the risk set, given survival up to time \(t\), and the
numerator is an adjusted survivor function that accounts for the
fraction of cured cases by time \(t\), which is \(1-\pi(t)\).

@@

In order to illustrate the package functionality, we examine a model of
coups d`etat in Belkin and Schofer (2003). Belkin and Schofer's paper
lends itself to re-analysis with a split-population duration model
because they explicitly distinguish long-term structural risk factors
for coups from more short-term triggering causes that can explain the
timing of a coup in an at-risk regime. They argue that many countries
never experience coups because coups are effectively impossible due to
structural factors, while others that never experience coups are
nevertheless at risk due to a different configuration of those same
factors. Using language which fits nicely with the class of models
described above, they write ``{[}t{]}riggers are not the source of the
original risk, and in the absence of structural causes, the presence of
triggering factors alone cannot lead to a coup. Hence, triggers should
not be equated with coup risk. Rather, they are factors that may
determine the exact timing of a coup in regimes that suffer from high
coup risk'' (Belkin and Schofer 2003, 598).

For their empirical test Belkin and Schofer develop a measure of
``structural coup risk'' which incorporates factors such as the strength
of democratic institutions and civil society, and a recent history of
successful coups. However, they implement a logistic regression model,
which does not capture the process as described in their quote above.
This is because the logistic regression model assumes that all
observations are at risk for a coup (i.e., the probability of a coup is
non-zero for all observations). Since their structural coup risk
indicator is developed precisely to distinguish between at risk and
``immune'' cases, the split-population model allows one to examine
whether the indicator effectively does so.

@@

\begin{CodeInput}
library("spduration")
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
                            freq = "year", ongoing = FALSE)
\end{CodeInput}

The \code{add\_duration} function takes as input a data frame with a
binary response variable--``coup''--that measures the occurrence of the
event, or failure, recorded over discrete time periods. We assume that
the data frame is in a cross-sectional time-series format that consists
of multiple observations over time for a number of subjects, in our case
countries.

Within the framework of duration modeling, these data conceptually
consist of ``spells'', which are repeated observations of a unit from
the time they enter the data (left-censoring) or after they experienced
the event of interest, until the next event or the end of observation
(right censoring). Table \ref{tab-ex} shows how a single country could
have two spells during the observation period. Observations for Portugal
make up two spells, one that begins in 1960, when our data start, to a
coup event in 1975, when a second spell starts and continues until
either the next coup or the data end. Canada on the other hand
experienced no coups and so observations for that country from 1960 to
2000 make up one spell.

\begin{table}
\begin{center}
\begin{tabular}{lrrrrr} \toprule
Country & Year & Coup & Spell ID & Duration \\ \midrule
Portugal & 1960 & 0 & 1 & 1  \\ 
\ldots & \ldots & 0 & 1 & \ldots \\
\ldots & 1975 & 1 & 1 & 16 \\ \midrule
Portugal & 1976 & 0 & 2 & 1 \\
\ldots & \ldots & 0 & 2 & \ldots \\
\ldots & 2000 & 0 & 2 & 20  \\ \midrule
Canada & 1960 & 0 & 3 & 1 \\
\ldots & \ldots & 0 & 3 & \ldots \\ 
\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
\end{tabular}
\end{center}
\caption{Example data frame with select duration variables.} \label{tab-ex}
\end{table}

The \code{add_duration} functions identifies spells in the data and
returns the data frame with several additional variables needed for
estimation. The most important among these are a counter of elapsed time
in a spell, ``duration''. It also codes that a spell was at risk
(``atrisk==1'') if it ended in a failure event, and not at risk
otherwise. These are then used as response variables for the formula in
the \code{spdur} function.

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data and will try to convert the
\code{tID} input to class ``Date'' given the provided argument in
\code{freq}. Another important question is how to handle consecutive 1's
in the response variable provided. This is controlled with the
\code{ongoing} argument. In some cases like civil war occurrence, the
\emph{y} variable records all years during which a country experienced a
civil war. Here \code{ongoing} should be set to true so that the models
try to predict the onset of the civil war, but disregard ongoing
conflicts (\code{failure} is set to \code{NA} for these cases, dropping
them from analysis). With ongoing set to false, successive 1's in
\emph{y} are treated as distinct and new failures, and kept in the
analysis. This makes sense for discrete, short-term events like coups.
Countries can experience distinct coups in successive years.

The \code{spdur} function is the primary function in the package and
produces a regression model object of class \code{spdur} which can then
be used with further methods. We begin by fitting first a Weibull and
then a log-logistic split-population duration model using the coups
data, including the measure of coup risk in the logit (risk) equation.

\begin{CodeChunk}
\begin{CodeInput}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = bscoup, silent = TRUE)

loglog_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = bscoup, distr = "loglog", silent = TRUE)
\end{CodeInput}
\end{CodeChunk}

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. These can be
exported by calling \code{xtable} directly on the ``spdur'' object to
produce Table 2.

\begin{CodeChunk}
\begin{CodeInput}
library("xtable")
tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
              label = "loglog_table")
print(tbl, caption.placement = "top", comment = FALSE, include.rownames = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
  milreg & -1.13 & 0.21 & 0.00 \\ 
  instab & -0.09 & 0.02 & 0.00 \\ 
  regconf & -2.52 & 2.16 & 0.24 \\ 
  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
  couprisk & 0.59 & 0.32 & 0.07 \\ 
  wealth & -0.36 & 0.28 & 0.20 \\ 
  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
  rwar & -0.52 & 0.94 & 0.58 \\ 
  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
  samerica & 2.09 & 1.45 & 0.15 \\ 
  camerica & -0.39 & 0.73 & 0.59 \\ 
   \hline
\end{tabular}
\caption{Coup model with log-logistic hazard.} \label{loglog_table}
\end{table}

Table \ref{loglog_table} shows estimates from the duration equation,
beginning with the intercept, and then estimates from the risk equation.
The duration component of the model is in accelerated failure time
format and the coefficient estimates are on the log of expected time to
failure. The negative coefficient for military regimes for example means
that the expected time to a coup is shorter in military regimes than
non-military regimes, holding all other factors constant. In the risk
equation, positive coefficients mean higher risk of coup. Thus, military
regimes have a higher risk of experiencing a coup.

One may also use the \code{AIC} and \code{BIC} function to calculate the
information criterion statistics for \code{spdur} objects. In our
example, both models are close enough in both statistics to be
indistinguishable, so we will continue to focus on the log-logistic
form.

\begin{CodeChunk}
\begin{CodeInput}
matrix(c(
  AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
  ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
\end{CodeInput}
\begin{CodeOutput}
             AIC      BIC
Weibull 1329.908 1348.972
Loglog  1331.214 1350.278
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
plot_hazard(loglog_model, main = "A")
\end{CodeInput}
\end{CodeChunk}

@@

As mean values are maybe not that interesting, we can also generate
scenarios by entering values manually. The values are used in the same
order that variables are specified in the equations used to estimate the
models, which also corresponds to the order of variables in Table 2.
Note the inclusion of an intercept term. This produces plot B in Figure
\ref{hazard-ex}

\begin{CodeChunk}
\begin{CodeInput}
plot_hazard(loglog_model, 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
\end{CodeInput}
\end{CodeChunk}

The hazard plots can also be called with
\code{plot(..., type = "hazard")}. Without a type argument,
\code{plot.spdur} will produce a separation plot, which is a graphical
display for evaluating model predictions (Greenhill, Ward, and Sacks
2011). The code below produces Figure \ref{insamp}.

@@

Underlying both plotting functions is the \code{predict} function, which
can be used on an object of class \code{spdur} to generate several kinds
of predictions, including the probability that an observation is
``at-risk'' and the probability of failure for a given time period.

\subsection{Out-of-sample testing}\label{out-of-sample-testing}

Finally, we demonstrate how to evaluate a model's out of sample
predictions. We will use data from 1996 onwards as the test set, and
prior data for training purposes. The \code{add_duration} function
retrospectively codes the risk variable based on how a spell ended, and
we therefore need to take care in how we add the duration variables for
each data set. For the training data we need to subset the training set
first, so that coups in the tests set don't influence the risk coding in
the training data.

\begin{CodeChunk}
\begin{CodeInput}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
                          freq = "year", ongoing = FALSE)
\end{CodeInput}
\end{CodeChunk}

For the test set it is recommended to add the duration variables and \textit{then} subset the test set. Since the test set is later in time than the training set we do not have to worry about contamination of the risk coding, but if we subset the data before building the duration variables we will start all duration counters at 1996, when in fact we can safely use the previous historic coup information. To do this we need to build the duration variables first:
